Towards Semantic Patches

### 2010-06-18 papers/MIPS2010/abstract.tex
* Un peu de texte en plus.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@103 403e655e-980a-47bd-8c60-1ee5348bb6b7

% Commandes à désactiver avant la soumission.
% Remarques sur la structure.
\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
% Remarques sur le contenu.
\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}

@@

With the previous remarks in mind, we designed a generic framework of semantic
patches that can be described using three layers which are summarized in 
Figure~\ref{fig:architecture}.

@@

The external layer offers views on the objects of the theory. A
view~$w$ is a human readable, and editable, partial representation of
an object~$t$. A view is obtained from an object through a
projection~$\pi$.  The user transforms the view $w$ into a view $w'$
which must be $\textit{put back}$ into the metatheory as a modified
object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.

\paragraph{From an abstract proof tree to a concrete view, and back}
\-

\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
  pas très pratique. Pourtant, c'est moralement sur cet objet 
  que s'opère le développement et non sur le code source, qui
  n'en est qu'une représentation. On justifie la nécessité de
  projection et on fait référence aux lentilles de Pierce pour
  expliquer $\sigma$ et $\pi$.}

@@

>>>>

% Commandes à désactiver avant la soumission.
% Remarques sur la structure.
%\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
% Remarques sur le contenu.
%\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}

\newcommand{\remplan}[1]{}
\newcommand{\remtext}[1]{}

@@

With the previous remarks in mind, we designed a generic framework of semantic
patches that can be described using three layers which are summarized in 
Figure~\ref{fig:architecture}. We now describe the role of each layer.

@@

The external layer offers views on the objects of the theory. A
view~$w$ is a human readable, and editable, partial representation of
an object~$t$. A view is obtained from an object through a
projection~$\pi$.  The user transforms the view $w$ into a view $w'$
which must be $\textit{put back}$ into the metatheory as a modified
object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.  In
concrete syntax, an object~$t$ may be denoted simply using its
syntactic definition but also, more interestingly, by a reference into
the repository or even as the application of a metatheorem.

The implementation of this architecture raises some interesting
technical difficulties that can be handled, but only partially, 
by existing technology. Indeed, even if internalization and 
externalization are easily implemented thanks to the expressivity
of the kernel language, the projection, putback, differenciation
and integration operations are more intricate to realize.

\paragraph{From an abstract proof tree to a concrete view, and back}
\-
\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
  pas très pratique. Pourtant, c'est moralement sur cet objet 
  que s'opère le développement et non sur le code source, qui
  n'en est qu'une représentation. On justifie la nécessité de
  projection et on fait référence aux lentilles de Pierce pour
  expliquer $\sigma$ et $\pi$.}

The problem of updating some data according to a modification that was
done on a partial view of it is a well-known issue in database
litterature called the \textit{view-update} problem. Pierce \textit{et
  al} have recently attacked it from a language design approach. Their
idea is to develop projection and putback functions simultaneously by
writing a so-called lens in a bidirectional programming
language~\cite{pierce-bidirectional}. Roughly speaking, a lense is a
software component that can be used to produce output from input but
also input from output (hence the bidirectionality). A lense is able
to keep track of the information that was erased by a projection to
produce a projected output from the initial input. When we supply a
modified output to the lense, it can elaborate, from the initial input
and under certain conditions, the missing part of the modified input.

We plan to design a bidirectional language to define the
projection~function~$\pi$ and the putback~function~$\sigma$ as a
lense. Yet, we cannot just import existing lenses combinators from the
Harmony system that already works on trees. Indeed, most of the time,
our trees have extra structural constraints induced by the presence of
binders.

@@

Computing the minimal edition script between two abstract syntax trees
is a NP-hard problem that has been heavily studied, in particular in
bioinformatics.  We guess that, in practice, we can restrict the
search to local transformations that preserve substructures (that
roughly consist in inserting, removing or moving around entire
subtrees). In that case, an existing polynomial algorithm might be
used~\cite{REF!}. Besides, abstract trees are relativily small in
practice.

We are more concerned about our ability to elaborate high-level patchs
from these low-level edition scripts. Actually, it is easy to deduce
from a low-level edition script, a low-level patch, based on object
syntax constructor elimination and introduction. However, such a
low-level patch may forbid a more efficient reuse of existing
objects. For instance, let us assume a transformation of an
object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
would represents the transformation as the replacement of the first
subterm $x$ by the subterm~$2$ followed by the replacament of the
second subterm $x$ by the subterm~$2$. If we had an object-level type
derivation for $x + x$, the integration function would have to build
two subderivations, one for each replaced subterms.  Instead, a
high-level patch would represent that transformation as a substitution
of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
that case, the integration function has the opportunity to apply a
metatheorem about the stability of typing derivations under
substitution and finally reuse the initial typing derivation
directly. Inferring automatically such a high-level patch seems an
unreachable objective~: there is clearly no most general high-level
patch. Nonetheless, an interactive process can be designed to help the
user building his high-level patch from a set of pre-computed
interpretations of the low-level patch.
